Supplemental Materials: \emph{Simultaneous segmentation and registration of diffusion MR images of the brain driven by active-contours}

### 2015-02-27 Publications/2015-NeuroImage/03-introduction.tex
paper version candidate. with abstract

\Gls*{dmri} data are usually acquired with \gls*{epi} schemes.
In order to quickly probe the diffusion process within the brain in many different orientations,
  these sequences minimize the bandwidth of the \gls*{pe} direction.
This low bandwith originates the artifactual displacement of locations presenting small deviations
  from the main magnetic field $B_0$.
The principal source of these deviations ($\Delta B_0$) is the discontinuity of magnetic
  susceptibility at tissue interfaces.
Tissue/air boundaries show large steps of susceptibility and this artifact is
  particularly evident in the surroundings of the orbito-frontal lobe or the temporal
  bone of both hemispheres.
The displacements field $U_s$ along the \gls*{pe} axis can be estimated theoretically
  \citep{jezzard_correction_1995}:

@@

where $\vec{r}$ is the true location in the undistorted domain as in \eqref{eq:transform},
  $\gamma$ is the gyromagnetic ratio, $T_{acq}$ is the readout time,
  $s_{PE}$ is the pixel spacing along \gls*{pe}, and $\hat{\vec{e}}_{PE}$ the unitary
  vector along \gls*{pe}.
Several methodologies have been proposed to estimate $U_s$ and fix the artifact.
The first solution was acquiring extra images to estimate $\Delta B_0$ using field mapping
  techniques \citep{jezzard_correction_1995,andersson_modeling_2001}.
A second set of methods aquired extra images reproducing the same \gls*{epi} scheme but switching
  the \gls*{pe} axis \citep{chiou_simple_2000} or reversing the direction of gradient
  increments in the same axis \citep{cordes_geometric_2000,holland_efficient_2010}.
\cite{kybic_unwarping_2000} proposed a correction using nonlinear registration of the \gls*{dmri}
  data versus a \gls*{t2} image.
Further developments \citep{irfanoglu_susceptibility_2011} extended this method to the
  three-dimensional problem.
All methods require extra \gls*{mr} aquisitions. Thus, a tipical protocol for the extraction
  of structural connectivity comprehends \gls*{dmri}, \gls*{t1}, and the acquisition demanded
  by the correction method of choice.

\emph{Regseg} is related to the methods based on nonlinear registration of the \emph{b0}
  image\footnote{The term \emph{b0} refers to one or more images within the
  \gls*{dmri} dataset that are acquired with a null or very low gradient intensity or
   \emph{b}-value used as signal reference.}
  to the \gls*{t2} image, since we also render the correction as a
  registration problem.
Conversely, the method does not require any of the previously mentioned extra acquisitions.
A fundamental antecedent to the present approach is found in \citep{greve_accurate_2009}.
They proposed a registration method to align contours extracted from the \gls*{t1}
  to \gls*{dmri} images using a linear transformation model.
Similarly, we exploit the precise surfaces that can be extracted with
  \emph{FreeSurfer}~\citep{fischl_freesurfer_2012},
  and then map this prior into warped data, in which the scale of the
  imaged structures is in the range of the image resolution or above.
Our approach differs from \citep{greve_accurate_2009} in two fundamental choices.
First, we use a nonlinear deformation model that enables the correction of the
  susceptibility-derived distortions \citep{jezzard_correction_1995}.
\cite{greve_accurate_2009} circumvented the nonlinearity by identifying tipically
  distorted regions and removing them from the metric computation.
Second, our matching criteria does not necessarily need for clear intensity
  steps at the tissue interfaces where surfaces must be fitted, thanks to the use of
  active contours without edges \citep{chan_active_2001}.
The hypothesis underpinning our research is that, in those situations,
  image registration can be reliably performed by searching for homogeneous
  regions in one image corresponding to a prior knowledge of the shape of objects in
  the moving (distorted) coordinate system.

>>>>

\Gls*{dmri} data are usually acquired with \gls*{epi} schemes.
In order to quickly probe the diffusion process within the brain in many different orientations,
  these sequences minimize the bandwidth of the \gls*{pe} direction.
This low bandwidth originates the artifactual displacement of locations presenting small deviations
  from the main magnetic field $B_0$.
The principal source of these deviations ($\Delta B_0$) is the discontinuity of magnetic
  susceptibility at tissue interfaces.
Tissue/air boundaries show large steps of susceptibility and this artifact is
  particularly evident in the surroundings of the orbito-frontal lobe or the temporal
  bone of both hemispheres.
The displacements field $U_s$ along the \gls*{pe} axis can be estimated theoretically
  \citep{jezzard_correction_1995}:

@@

where $\vec{r}$ is the true location in the undistorted domain as in \eqref{eq:transform},
  $\gamma$ is the gyromagnetic ratio, $T_{acq}$ is the readout time,
  $s_{PE}$ is the pixel spacing along \gls*{pe}, and $\hat{\vec{e}}_{PE}$ the unitary
  vector along \gls*{pe}.
Several methodologies have been proposed to estimate $U_s$ and fix the artifact.
The first solution was acquiring extra images to estimate $\Delta B_0$ using field mapping
  techniques \citep{jezzard_correction_1995,andersson_modeling_2001}.
A second set of methods acquired extra images reproducing the same \gls*{epi} scheme but switching
  the \gls*{pe} axis \citep{chiou_simple_2000} or reversing the direction of gradient
  increments in the same axis \citep{cordes_geometric_2000,holland_efficient_2010}.
\cite{kybic_unwarping_2000} proposed a correction using nonlinear registration of the \gls*{dmri}
  data versus a \gls*{t2} image.
Further developments \citep{irfanoglu_susceptibility_2011} extended this method to the
  three-dimensional problem.
All methods require extra \gls*{mr} acquisitions. Thus, a typical protocol for the extraction
  of structural connectivity comprehends \gls*{dmri}, \gls*{t1}, and the acquisition demanded
  by the correction method of choice.

\emph{Regseg} is related to the methods based on nonlinear registration of the \emph{b0}
  image\footnote{The term \emph{b0} refers to one or more images within the
  \gls*{dmri} dataset that are acquired with a null or very low gradient intensity or
   \emph{b}-value used as signal reference.}
  to the \gls*{t2} image, since we also render the correction as a
  registration problem.
Conversely, the method does not require any of the previously mentioned extra acquisitions.
A fundamental antecedent to the present approach is found in \citep{greve_accurate_2009}.
They proposed a registration method to align contours extracted from the \gls*{t1}
  to \gls*{dmri} images using a linear transformation model.
Similarly, we exploit the precise surfaces that can be extracted with
  \emph{FreeSurfer}~\citep{fischl_freesurfer_2012},
  and then map this prior into warped data, in which the scale of the
  imaged structures is in the range of the image resolution or above.
Our approach differs from \citep{greve_accurate_2009} in two fundamental choices.
First, we use a nonlinear deformation model that enables the correction of the
  susceptibility-derived distortions \citep{jezzard_correction_1995}.
\cite{greve_accurate_2009} circumvented the nonlinearity by identifying typically
  distorted regions and removing them from the metric computation.
Second, our matching criteria does not necessarily need for clear intensity
  steps at the tissue interfaces where surfaces must be fitted, thanks to the use of
  active contours without edges \citep{chan_active_2001}.
The hypothesis underpinning our research is that, in those situations,
  image registration can be reliably performed by searching for homogeneous
  regions in one image corresponding to a prior knowledge of the shape of objects in
  the moving (distorted) coordinate system.

### 2015-02-27 Publications/2015-NeuroImage/04-methods.tex
paper version candidate. with abstract

An early ``morphing''-segmentation approach was proposed in \citep{bertalmio_morphing_2000}.
Shortly, \cite{yezzi_variational_2001} presented the first method including a full solution to
  the registration, where the energy functional is defined simultaneously in the target
  and reference images with an affine transformation supporting the coordinates mapping.
\cite{vemuri_joint_2003} proposed an atlas-based registration using a level-sets
  framework and only one \gls*{pde} for first time.
The principal application of these techniques was to perform registration between timesteps in
  time-series or between different images \citep{paragios_level_2003}.
\cite{unal_coupled_2005} and later \cite{wang_joint_2006},
  extended the ``two \glspl*{pde}'' method of \cite{yezzi_variational_2001}
  to nonlinear registration with a free deformation field.
\cite{droske_mumfordshah_2009} reviewed the latter set of techniques, and proposed two different
  approaches to apply the Mumford-Shah functional \citep{mumford_optimal_1989} in simultaneous
  registration-segmentation, through the propagation of the deformation field from
  the contours to the whole image definition.
Finally, \cite{gorthi_active_2011} presented a comprehensive generalization of the
  existing methodologies based on level sets in the application of atlas-based segmentation.
Recently, \cite{guyader_combined_2011} proposed a simultaneous segmentation and
  registration method close to the one proposed here.
Our method differs on two key features:
  it allows multiple active contours for segmentation, and the regularization strategy.
Whereas \cite{guyader_combined_2011} proposed a nonlinear elasticity smoother, we rely on
  the inherent smoothness of the deformation field, and on the anisotropic regularization
  proposed by \cite{nagel_investigation_1986}.

@@

\paragraph*{Settings, implementation details, and convergence}
\label{sec:conv_report}
The registration parameters (such as $\tau$, the B-Spline grid resolutions,
  target image smoothing, etc.) and implementation details (for instance the sparse matrix approach
  to fast interpolation) are dicussed in the \suppl{}, section S1.
The actual choices of parameter settings are publicly distributed with the source code of the experiments.
Additionally, we release our software along with a tool for generating convergence reports to
  demonstrate the behavior of \emph{regseg} and help scientists configure their own experiments.
One sample report is found in the \suppl{}, section S1.3.

>>>>

An early ``morphing''-segmentation approach was proposed in \citep{bertalmio_morphing_2000}.
Shortly, \cite{yezzi_variational_2001} presented the first method including a full solution to
  the registration, where the energy functional is defined simultaneously in the target
  and reference images with an affine transformation supporting the coordinates mapping.
\cite{vemuri_joint_2003} proposed an atlas-based registration using a level-sets
  framework and only one \gls*{pde} for first time.
The principal application of these techniques was to perform registration between time steps in
  time series or between different images \citep{paragios_level_2003}.
\cite{unal_coupled_2005} and later \cite{wang_joint_2006},
  extended the ``two \glspl*{pde}'' method of \cite{yezzi_variational_2001}
  to nonlinear registration with a free deformation field.
\cite{droske_mumfordshah_2009} reviewed the latter set of techniques, and proposed two different
  approaches to apply the Mumford-Shah functional \citep{mumford_optimal_1989} in simultaneous
  registration-segmentation, through the propagation of the deformation field from
  the contours to the whole image definition.
Finally, \cite{gorthi_active_2011} presented a comprehensive generalization of the
  existing methodologies based on level sets in the application of atlas-based segmentation.
Recently, \cite{guyader_combined_2011} proposed a simultaneous segmentation and
  registration method close to the one proposed here.
Our method differs on two key features:
  it allows multiple active contours for segmentation, and the regularization strategy.
Whereas \cite{guyader_combined_2011} proposed a nonlinear elasticity smoother, we rely on
  the inherent smoothness of the deformation field, and on the anisotropic regularization
  proposed by \cite{nagel_investigation_1986}.

@@

\todo[inline]{Include here the final update equation using the FT, introducing the
step size $\tau$, and refer to the supplementals for intermediate steps}

\paragraph*{Settings, implementation details, and convergence}
\label{sec:conv_report}
The registration parameters (such as $\tau$, the B-Spline grid resolutions,
  target image smoothing, etc.) and implementation details (for instance the sparse matrix approach
  to fast interpolation) are discussed in the \suppl{}, section S1.
The actual choices of parameter settings are publicly distributed with the source code of the experiments.
Additionally, we release our software along with a tool for generating convergence reports to
  demonstrate the behavior of \emph{regseg} and help scientists configure their own experiments.
One sample report is found in the \suppl{}, section S1.3.

### 2015-02-27 Publications/2015-NeuroImage/05-results.tex
paper version candidate. with abstract

A total of 1200 registration experiments (4 phantom types $\times$ 150 random warpings
  $\times$ 2 resolutions) were run.
Parameter settings of every processing step in the experiments are made available along with
  the source code.
For each registration, the averaged Hausdorff distance (see \nameref{sec:experiments_evaluation})
  of the inner and the outer surfaces was computed.
The results showed a consistent and high accuracy, below voxel resolution.
\autoref{fig:phantom} (block C) presents the box plots by model type corresponding
  to the two sets of resolutions of generated phantoms.
To support that the missregistration errors averaged per experiment was significantly
  under the resolution of the target image, we proceeded as follows.
First, we confirmed that the error distributions were skewed using a Shapiro-Wilk test of
  normality.
All the distributions of errors under test (4 phantom types $\times$ 2 resolutions) resulted
  non-normal with $p<0.001$.
Consequently, we used the non-parametric Wilcoxon signed-rank test along with Bonferroni
  correction for multiple comparisons ($N=150$).
Averaged errors resulted significantly lower than voxel resolution with $p < (0.001 / 150)$
  for all the tests (4 phantom types $\times$ 2 resolutions).
Since statistical tests may not be conclusive enough, we also computed the confidence intervals
  (\autoref{tab:ci_phantom}).

@@

\paragraph*{Assessing the segmentation model}\label{sec:res_model_and_metric} %
%
Preliminarly, we investigated the aptness of our segmentation model to play as cost function
  driving the registration process.
First, we evaluated several alternative models to segment the target bivariate images
  (\gls*{fa} and \gls*{adc}).
Based on the observed joint distribution of each (\suppl{}, {\color{red} section S2}),
  we selected the 6-regions scheme described in \nameref{sec:human_connectome}.

The distortions follow expression \eqref{eq:fieldmap} and hence, we easily reduced the
  error space to one dimension as follows.
The energy functional \eqref{eq:energy} was evaluated for the warpings given by
  a lineralized space of distortions
  $\hat{U} = \epsilon \cdot U_{true} = \vec{r} + \epsilon \cdot u(\vec{r})$,
  with $\epsilon \in [-1.1, 1.1]$.
The results of this experiment are shown in \autoref{fig:energymap}.
The metric consistently displayed its minimum at $\epsilon=0.0$ (ground-truth location).

@@

\begin{table}
		\centering
		\footnotesize
		\tabcolsep=0.05cm
    \begin{tabular}{L{1.8cm}ccc}
                            & Surf. 1  & Surf. 2 & Surf. 3 \\
    \hline
    CI \emph{regseg}        & 0.39 - 0.64 & 0.64 - 0.83 & 0.76 - 0.93 \\
    CI competing            & 2.36 - 3.37 & 2.65 - 3.19 & 2.82 - 3.40 \\
    \hline
    ANOVA (f-stat/p-value)  & (51.81/1.5e-06) & (148.25/8.1e-10) & (143.01/1.1e-09) \\
    \hline
    \end{tabular}
    \caption{All the 95\% CIs of the \gls*{swindex} distributions computed for the
      surfaces of interest were lower for \emph{regseg} with respect to the competing
      method.
    CIs reported in this table were computed using bootstrapping with mean as location
      statistic and $10^4$ samples.
    One-way-ANOVA tests indicated that there is a significant difference between our method and
      the competing method of choice.
    }\label{tab:results_real}
\end{table}

\begin{figure*}
	\includegraphics[width=\linewidth]{figures/figure05.pdf}
	\caption{The evaluation instrument automatically renders the overlay of the
	  target \gls*{fa} image and the resulting contours after distortion correction (yellow color) and
	  their theoretical location using the ground-truth (green color).
	The first two rows show several axial slices for \emph{regseg} (``proposed'' row) and the
	  competing method.
	The last two rows represent the sagittal view.
	We intentionally ommited the coronal slicing due to it is the less informative given the properties
	  of the distortion.
	Red arrows point to regions where the accuracy of the proposed method more clearly overperformed
	  the competing method.
	}\label{fig:results_real}
\end{figure*}

>>>>

A total of 1200 registration experiments (4 phantom types $\times$ 150 random warpings
  $\times$ 2 resolutions) were run.
Parameter settings of every processing step in the experiments are made available along with
  the source code.
For each registration, the averaged Hausdorff distance (see \nameref{sec:experiments_evaluation})
  of the inner and the outer surfaces was computed.
The results showed a consistent and high accuracy, below voxel resolution.
\autoref{fig:phantom} (block C) presents the box plots by model type corresponding
  to the two sets of resolutions of generated phantoms.
To support that the misregistration errors averaged per experiment was significantly
  under the resolution of the target image, we proceeded as follows.
First, we confirmed that the error distributions were skewed using a Shapiro-Wilk test of
  normality.
All the distributions of errors under test (4 phantom types $\times$ 2 resolutions) resulted
  non-normal with $p<0.001$.
Consequently, we used the non-parametric Wilcoxon signed-rank test along with Bonferroni
  correction for multiple comparisons ($N=150$).
Averaged errors resulted significantly lower than voxel resolution with $p < (0.001 / 150)$
  for all the tests (4 phantom types $\times$ 2 resolutions).
Since statistical tests may not be conclusive enough, we also computed the confidence intervals
  (\autoref{tab:ci_phantom}).

@@

\paragraph*{Assessing the segmentation model}\label{sec:res_model_and_metric} %
%
Preliminarily, we investigated the aptness of our segmentation model to play as cost function
  driving the registration process.
First, we evaluated several alternative models to segment the target bivariate images
  (\gls*{fa} and \gls*{adc}).
Based on the observed joint distribution of each (\suppl{}, {\color{red} section S2}),
  we selected the 6-regions scheme described in \nameref{sec:human_connectome}.

The distortions follow expression \eqref{eq:fieldmap} and hence, we easily reduced the
  error space to one dimension as follows.
The energy functional \eqref{eq:energy} was evaluated for the warpings given by
  a linearized space of distortions
  $\hat{U} = \epsilon \cdot U_{true} = \vec{r} + \epsilon \cdot u(\vec{r})$,
  with $\epsilon \in [-1.1, 1.1]$.
The results of this experiment are shown in \autoref{fig:energymap}.
The metric consistently displayed its minimum at $\epsilon=0.0$ (ground-truth location).

@@

\begin{table}
		\centering
		\footnotesize
		\tabcolsep=0.05cm
    \begin{tabular}{L{1.8cm}ccc}
                            & $\Gamma_{VdGM}$  & $\Gamma_{WM}$ & $\Gamma_{pial}$ \\
    \hline
    CI \emph{regseg}        & 0.39 - 0.64 & 0.64 - 0.83 & 0.76 - 0.93 \\
    CI competing            & 2.36 - 3.37 & 2.65 - 3.19 & 2.82 - 3.40 \\
    \hline
    ANOVA (f-stat/p-value)  & (51.81/1.5e-06) & (148.25/8.1e-10) & (143.01/1.1e-09) \\
    \hline
    \end{tabular}
    \caption{All the 95\% CIs of the \gls*{swindex} distributions computed for the
      surfaces of interest were lower for \emph{regseg} with respect to the competing
      method.
    CIs reported in this table were computed using bootstrapping with mean as location
      statistic and $10^4$ samples.
    One-way-ANOVA tests indicated that there is a significant difference between our method and
      the competing method of choice.
    }\label{tab:results_real}
\end{table}

\begin{figure*}
	\includegraphics[width=\linewidth]{figures/figure05.pdf}
	\caption{The evaluation instrument automatically renders the overlay of the
	  target \gls*{fa} image and the resulting contours after distortion estimation (yellow color) and
	  their theoretical location using the ground-truth (green color).
	The first two rows show several axial slices for \emph{regseg} (``proposed'' row) and the
	  competing method.
	The last two rows represent the sagittal view.
	We intentionally omitted the coronal slicing due to it is the less informative given the properties
	  of the distortion.
	Red arrows point to regions where the accuracy of the proposed method more clearly overperformed
	  the competing method.
	\todo[inline]{This figure will also have the box-plots of the errors of surfaces for both regseg
	and the t2b method}
	}\label{fig:results_real}
\end{figure*}
